---
title: "Analysis of Heart Failure Risk Associated with Sodium and Protein Consumption"
subtitle: "NHANES Data (Combined Cycles 2011-2018)"
author: "João Silva, Marco Pereira, Mariana Ribeiro"
date: "`r Sys.Date()`"
output: pdf_document
---

# Abstract

This study examines the association between habitual dietary sodium and protein intake and the self-reported prevalence of heart failure (HF) in the US adult population, adjusting for socioeconomic determinants and metabolic comorbidities. Data from four combined cycles of the National Health and Nutrition Examination Survey (NHANES 2011–2018) were analysed using complex survey-weighted logistic regression. Nutritional exposures were energy-adjusted using Willett's residual method. The sample includes adults aged 18 and older with complete covariate data. Results are reported as weighted Odds Ratios (OR) with 95% confidence intervals, representative of the non-institutionalised US adult population.

# 1. Preparation and Package Loading

In this section, we load the necessary libraries for API extraction, data manipulation (`dplyr`, `purrr`), complex survey analysis (`survey`), descriptive statistics (`tableone`), model diagnostics (`pROC`, `car`), and visualisation (`ggplot2`). We also include `knitr` and `kableExtra` for rendering elegant tables in the final output.

```{r packages, echo=TRUE}
# Install if necessary:
# install.packages(c("dplyr", "survey", "tableone", "knitr", "kableExtra",
#                    "haven", "nhanesA", "purrr", "tidyr", "ggplot2", "pROC", "car"))
library(haven)
library(nhanesA)
library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(survey)
library(stringr)
library(tableone)
library(knitr)
library(kableExtra)
library(ggplot2)
library(pROC)
library(car)
```

# 2. Data Import (NHANES Interface)

The following block mirrors the original mapping and extraction from the CDC database for cycles G to J (2011-2018). Combining four survey cycles increases the statistical power of the analysis.

**Methodological Note — Complete Case Analysis:** Missing data (NAs) in clinical and socioeconomic covariates (e.g., income ratio, BMI, blood pressure) cause listwise deletion during logistic regression. If NAs enter the survey design object, the model silently discards incomplete records, distorting sample weights. A Complete Case Analysis (CCA) filter is therefore applied explicitly, ensuring the statistical model operates on fully observed data only. The number of participants excluded by CCA is reported transparently in Section 2.1.

> **Note:** This extraction block is set to `eval=FALSE` to avoid live API calls on every compilation. It assumes `data/clean_dataset.rds` has already been generated by running this chunk once interactively. To regenerate the dataset, set `eval=TRUE` and ensure network access to the CDC servers.

```{r data_import, eval=FALSE, echo=TRUE}
cycles <- c("G", "H", "I", "J") 

get_clinical_data <- function(suffix) {
  
  # Table Extraction
  demo <- nhanes(paste0("DEMO_", suffix))    # Demographics (PIR, Gender, Age, Weights)
  mcq  <- nhanes(paste0("MCQ_", suffix))     # Medical Conditions (Heart Failure)
  nutr <- nhanes(paste0("DR1TOT_", suffix))  # Total Nutrition (Sodium, Protein)
  bmx  <- nhanes(paste0("BMX_", suffix))     # Body measures (BMI)
  bpx  <- nhanes(paste0("BPX_", suffix))     # Blood Pressure (Measured)
  diq  <- nhanes(paste0("DIQ_", suffix))     # Diabetes
  pro  <- nhanes(paste0("DR1IFF_", suffix))  # Individual Foods
  med  <- nhanes(paste0("RXQ_RX_", suffix))  # Prescription Medications 

  # Cleaning and Selection
  # --- Demographics ---
  d_clean <- demo %>%
    select(SEQN, Age = RIDAGEYR, Gender = RIAGENDR,
           Poverty_Index_Ratio = INDFMPIR, Education = DMDEDUC2,
           Weight_MEC = WTMEC2YR, Strata = SDMVSTRA, PSU = SDMVPSU)
  
  # --- Heart Failure outcome ---
  m_clean <- mcq %>% select(SEQN, HeartFailure = MCQ160B)
  
  # --- Nutrition ---
  nutr_clean <- nutr %>%
    select(SEQN, Sodium_mg = DR1TSODI, Protein_g = DR1TPROT,
           Energy_kcal = DR1TKCAL)

  # --- BMI ---
  b_clean <- bmx %>% select(SEQN, BMI = BMXBMI)
  
  # --- Diabetes ---
  diq_clean <- diq %>% select(SEQN, Diabetes = DIQ010)
  
  # --- Blood Pressure: mean of up to 3 readings ---
  bpx_clean <- bpx %>%
    rowwise() %>%
    mutate(
      Mean_Systolic  = mean(c(BPXSY1, BPXSY2, BPXSY3),  na.rm = TRUE),
      Mean_Diastolic = mean(c(BPXDI1, BPXDI2, BPXDI3),  na.rm = TRUE)
    ) %>%
    select(SEQN, Mean_Systolic, Mean_Diastolic)
  
  # --- Protein origin: animal vs. plant via USDA FNDDS first-digit food group ---
  # First digit 1-3 = animal-origin groups (dairy/eggs, meat, fish/seafood)
  # First digit 4-7 = plant-origin groups (legumes, grains, fruits, vegetables)
  # NOTE: This is an approximation. Mixed/processed foods may be misclassified.
  # The limitation is acknowledged in Section 7 (Limitations).
  pro_clean <- pro %>%
    group_by(SEQN) %>%
    summarise(
      Total_Animal_Protein_g = sum(
        if_else(substr(as.character(DR1IFDCD), 1, 1) %in% c("1","2","3"),
                as.numeric(DR1IPROT), 0),
        na.rm = TRUE),
      Total_Plant_Protein_g = sum(
        if_else(substr(as.character(DR1IFDCD), 1, 1) %in% c("4","5","6","7"),
                as.numeric(DR1IPROT), 0),
        na.rm = TRUE),
      .groups = "drop"
    )
  
  # --- Medication --- TO BE CHANGED
  hf_icd_codes <- c("I50", "I50.9", "I50.1", "I50.2", "I50.3", "I50.4", "I50.9P")
  
  med_clean <- med %>%
        rename_with(toupper) %>% group_by(SEQN) %>%
        summarise(
                  HF_Medication = as.factor(any(
                        if_any(any_of(c("RXDRSC1", "RXDRSC2", "RXDRSC3")),
                        ~ trimws(.x) %in% hf_icd_codes), na.rm = TRUE)),
                  .groups = "drop"
  )
 
  # --- Merge all tables on SEQN ---
  tabular_data_NHANES <- d_clean %>%
    left_join(m_clean,    by = "SEQN") %>%
    left_join(nutr_clean, by = "SEQN") %>%
    left_join(b_clean,    by = "SEQN") %>%
    left_join(diq_clean,  by = "SEQN") %>%
    left_join(bpx_clean,  by = "SEQN") %>%
    left_join(pro_clean,  by = "SEQN") %>%
    left_join(med_clean,  by = "SEQN") %>%
    mutate(Cycle = suffix)

  return(tabular_data_NHANES)
}

# Download and combine all four cycles
all_clinical <- map_df(cycles, get_clinical_data)

# Restrict to adults aged 18+
target_population <- all_clinical %>% filter(Age >= 18)

# Complete Case Analysis — explicit filter before survey design declaration
df_clean <- target_population %>%
  filter(
    !is.na(Sodium_mg), !is.na(Protein_g), !is.na(Energy_kcal),
    !is.na(Poverty_Index_Ratio), !is.na(BMI), !is.na(Mean_Systolic),
    !is.na(Total_Animal_Protein_g),
    HeartFailure %in% c("Yes", "No"),
    Diabetes     %in% c("Yes", "No", "Borderline"),
    !(Education  %in% c("Refused", "Don't Know", "Don't know"))
  )

# Save for downstream chunks
dir.create("data", showWarnings = FALSE)
saveRDS(df_clean,          file = "data/clean_dataset.rds")
saveRDS(target_population, file = "data/target_population.rds")  # kept for CCA report
```

```{r load_data}
# Load the pre-cleaned dataset for all downstream analysis
df_clean          <- readRDS("data/clean_dataset.rds")
target_population <- readRDS("data/target_population.rds")
```

## 2.1 Complete Case Analysis — Attrition Report

Transparent reporting of sample attrition is required to assess whether the CCA assumption (data missing at random) is tenable.

```{r cca_report}
n_adults  <- nrow(target_population)
n_final   <- nrow(df_clean)
n_dropped <- n_adults - n_final
pct_kept  <- round(n_final  / n_adults * 100, 1)
pct_lost  <- round(n_dropped / n_adults * 100, 1)

attrition_table <- data.frame(
  Stage = c(
    "Adults aged ≥18 (all cycles combined)",
    "Excluded: missing nutritional data",
    "Excluded: missing socioeconomic / clinical covariates",
    "Excluded: invalid / refused HF, diabetes, or education response",
    "Final analytic sample (complete cases)"
  ),
  N = c(
    n_adults,
    sum(is.na(target_population$Sodium_mg) | is.na(target_population$Protein_g) | is.na(target_population$Energy_kcal)),
    sum(!is.na(target_population$Sodium_mg) & (is.na(target_population$Poverty_Index_Ratio) | is.na(target_population$BMI) | is.na(target_population$Mean_Systolic))),
    sum(!is.na(target_population$Sodium_mg) & !(target_population$HeartFailure %in% c("Yes","No")), na.rm = TRUE),
    n_final
  )
)

kable(attrition_table, col.names = c("Stage", "n"), align = "lc",
      caption = "Sample attrition through complete case analysis.") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)

cat(sprintf(
  "\nRetained: %d of %d adults (%.1f%%). Excluded: %d (%.1f%%).\n",
  n_final, n_adults, pct_kept, n_dropped, pct_lost
))
```

> **Limitation note:** CCA assumes data are Missing At Random (MAR). Lower-income participants systematically have higher rates of missing dietary recall and clinical data, so the analytic sample may over-represent higher socioeconomic strata. This is acknowledged in Section 7.

# 3. Data Transformation and Methodologies

## 3.1 Cleaning and Clinical Factors

The dependent variable (Heart Failure) is converted into a strictly binary format to enable logistic regression. Categorical covariates are structured into factors with pre-defined reference categories

**Key methodological decisions:**

-   **Weight adjustment for pooled cycles (`WTMEC8YR`):** The original NHANES weight (`WTMEC2YR`) represents the US population over a 2-year cycle. Combining 4 cycles requires dividing by 4, per CDC guidelines for multi-cycle pooling. Failure to do so would artificially inflate the represented population to \>1.2 billion, distorting variance and confidence intervals.

-   **Diabetes categorisation:** Diabetes is assessed across 3 clinical levels — "No", "Pre-diabetes" (Borderline), and "Yes" — to prevent contamination of the healthy reference group with insulin-resistant individuals.

-   **Protein source thresholds:** Diet types are categorised using evidence-based thresholds from Perraud et al. (2023; DOI: 10.3389/fnut.2023.1178121). The "Mixed" reference category spans 30–75% animal protein, reflecting a clinically balanced omnivore diet.

```{r data_transformation}
analysis_data <- df_clean %>%
  mutate(
    # Outcome variable (binary)
    HeartFailure = ifelse(HeartFailure == "Yes", 1, 0),

    # Survey weights adjusted for 4-cycle pooling (8 years total)
    Weight_MEC8 = Weight_MEC / 4,

    # Education (ordered factor, reference: < 9th Grade)
    Education = factor(case_when(
      str_detect(Education, "Less than 9th")       ~ "< 9th Grade",
      str_detect(Education, "9-11th grade")         ~ "9th to 11th Grade",
      str_detect(Education, "High school graduate") ~ "High School Graduate",
      str_detect(Education, "Some college")         ~ "Some College",
      str_detect(Education, "College graduate")     ~ "College Graduate or +",
      TRUE ~ "Unknown"
    ), levels = c("< 9th Grade", "9th to 11th Grade",
                  "High School Graduate", "Some College",
                  "College Graduate or +")),

    # Diabetes (3-level, reference: No)
    Diabetes = factor(case_when(
      Diabetes == "No"         ~ "No",
      Diabetes == "Borderline" ~ "Pre-diabetes",
      Diabetes == "Yes"        ~ "Yes"
    ), levels = c("No", "Pre-diabetes", "Yes")),

    # Medication use: antihypertensive / diuretic flag
    Medication_Use = factor(
      ifelse(HF_Medication == TRUE, "Yes", "No"),
      levels = c("No", "Yes")
    ),

    # Protein origin ratio and categorical diet type
    Animal_Ratio = Total_Animal_Protein_g / Protein_g,
    Protein_Origin = factor(case_when(
      Animal_Ratio > 0.75 ~ "Animal",
      Animal_Ratio < 0.30 ~ "Plant",
      TRUE                ~ "Mixed"
    ), levels = c("Mixed", "Animal", "Plant"))
  )


```

## 3.2 Willett's Residual Method (Survey-Weighted)

To mitigate "volume bias", we adjust sodium and protein intakes for total energy intake. The goal is to identify individuals consuming a *disproportionately* high amount of sodium relative to their caloric needs, rather than penalising those with naturally high energy requirements (e.g., physically active individuals).

The residual regressions are fitted using `svyglm` to respect the complex survey design. This ensures the energy–nutrient relationship is estimated at the population level, not merely the sample level. The population mean is added back to residuals to preserve interpretability on the original measurement scale.

```{r willett}
# Temporary survey design object for weighted residual regression
nhanes_design_willett <- svydesign(
  id      = ~PSU,
  strata  = ~Strata,
  weights = ~Weight_MEC8,
  nest    = TRUE,
  data    = analysis_data
)

# Survey-weighted linear models for Willett residuals
model_sodium_svy <- svyglm(Sodium_mg ~ Energy_kcal, design = nhanes_design_willett)
model_prot_svy   <- svyglm(Protein_g ~ Energy_kcal, design = nhanes_design_willett)

# Population means (weighted) for re-centering residuals
mean_sodium_wtd <- coef(svymean(~Sodium_mg, nhanes_design_willett))
mean_prot_wtd   <- coef(svymean(~Protein_g, nhanes_design_willett))

# Append Willett-adjusted variables
analysis_data <- analysis_data %>%
  mutate(
    Sodium_Willett  = residuals(model_sodium_svy) + mean_sodium_wtd,
    Protein_Willett = residuals(model_prot_svy)   + mean_prot_wtd
  )
```

# 4. Weighted Analysis (Survey Design)

The `survey` package declares Primary Sampling Units (PSUs), strata, and sample weights. This overrides standard assumptions of equal-probability sampling, ensuring all inferences reflect the non-institutionalised US adult population.

```{r survey_design}
nhanes_design <- svydesign(
        id      = ~PSU, 
        strata  = ~Strata, 
        weights = ~Weight_MEC8, 
        nest    = TRUE, 
        data    = analysis_data
)
```

## Table 1: Demographic and Clinical Characteristics

Distribution of population characteristics stratified by heart failure status, adjusted to NHANES survey weights. P-values are computed using survey-appropriate tests (`svyttest` for continuous variables, `svychisq` for categorical variables)

```{r table_1}
vars_table <- c("Sodium_Willett", "Protein_Willett", "Protein_Origin",
                "Age", "Gender", "Poverty_Index_Ratio",
                "Education", "BMI", "Diabetes",
                "Mean_Systolic", "Medication_Use")

vars_cat <- c("Protein_Origin", "Gender", "Education",
              "Diabetes", "Medication_Use")

table_1_weighted <- svyCreateTableOne(
  vars       = vars_table,
  strata     = "HeartFailure",
  data       = nhanes_design,
  factorVars = vars_cat,
  test       = TRUE   # survey-appropriate tests (svyttest / svychisq)
)

kableone(
  table_1_weighted,
  showAllLevels  = TRUE,
  formatOptions  = list(big.mark = ","),
  caption        = "Table 1. Weighted demographic and clinical characteristics by heart failure status. Values are weighted means (SD) or weighted proportions (%). P-values from survey-adjusted tests."
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  footnote(
    general = "HF = Heart Failure; PIR = Poverty-Income Ratio; BMI = Body Mass Index. Sodium and Protein are energy-adjusted Willett residuals. Medication_Use flags antihypertensive or diuretic therapy.",
    general_title = "Note: "
  )
```

# 5. Logistic Model Results

A quasi-binomial model, appropriate for complex survey-weighted data with a binary outcome, is fitted using `svyglm`. Quasi-binomial correctly handles overdispersion relative to standard binomial and is the recommended family for this design.

```{r model}
final_model <- svyglm(
  HeartFailure ~ Sodium_Willett + Protein_Willett + Protein_Origin +
    Age + Gender + Poverty_Index_Ratio + Education +
    BMI + Diabetes + Mean_Systolic + Medication_Use,
  design = nhanes_design,
  family = quasibinomial()
)

# Extract the model summary to obtain the p-values
model_summary <- summary(final_model)$coefficients

# Odds Ratios, 95% CIs, and P-values
table_OR <- data.frame(
  Variable  = names(coef(final_model)),
  OR        = exp(coef(final_model)),
  CI_2.5    = exp(confint(final_model))[, 1],
  CI_97.5   = exp(confint(final_model))[, 2],
  P_Value   = model_summary[, 4], # The 4th column contains the Pr(>|t|) p-value
  row.names = NULL
) %>%
  mutate(
    # Format the P-value clinically (values below 0.05 become "< 0.05", otherwise show the number)
    P_Value_Str = case_when(
      P_Value < 0.05 ~ "< 0.05",
      TRUE ~ as.character(round(P_Value, 3))
    )
  ) %>%
  # Round only the OR and CI columns to 3 decimal places
  mutate(across(c(OR, CI_2.5, CI_97.5), ~ round(.x, 3))) %>%
  # Reorganise the columns and drop the raw numeric P_Value (keeping only the formatted one)
  select(Variable, OR, CI_2.5, CI_97.5, P_Value = P_Value_Str)

# Human-readable variable labels
table_OR <- table_OR %>%
  mutate(Variable = case_when(
    Variable == "(Intercept)"                        ~ "Intercept",
    Variable == "Sodium_Willett"                     ~ "Sodium: energy-adjusted (mg)",
    Variable == "Protein_Willett"                    ~ "Protein: energy-adjusted (g)",
    Variable == "Protein_OriginAnimal"               ~ "Animal protein (Ref: Mixed)",
    Variable == "Protein_OriginPlant"                ~ "Plant protein (Ref: Mixed)",
    Variable == "Age"                                ~ "Age (years)",
    Variable == "Gender_CleanFemale"                 ~ "Gender: Female (Ref: Male)",
    Variable == "Poverty_Index_Ratio"                ~ "Income-to-poverty ratio (PIR)",
    str_detect(Variable, "Education_Clean9th")       ~ "Education: 9th–11th grade",
    str_detect(Variable, "Education_CleanHigh")      ~ "Education: High school graduate",
    str_detect(Variable, "Education_CleanSome")      ~ "Education: Some college",
    str_detect(Variable, "Education_CleanCollege")   ~ "Education: College graduate or above",
    Variable == "BMI"                                ~ "Body mass index (BMI)",
    Variable == "Diabetes_CleanPre-diabetes"         ~ "Comorbidity: Pre-diabetes (Ref: No)",
    Variable == "Diabetes_CleanYes"                  ~ "Comorbidity: Diabetes (Ref: No)",
    Variable == "Mean_Systolic"                      ~ "Mean systolic blood pressure (mmHg)",
    Variable == "Medication_UseYes"                  ~ "Antihypertensive / diuretic use (Yes)",
    TRUE ~ Variable
  ))

kable(
  table_OR,
  col.names = c("Variable", "Odds Ratio", "95% CI (Lower)", "95% CI (Upper)", "P-Value"),
  align     = "lcccc",
  caption   = "Table 2. Adjusted Odds Ratios for heart failure from survey-weighted quasi-binomial logistic regression. Reference categories in parentheses. NHANES 2011–2018."
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  footnote(
    general       = "OR = Odds Ratio; CI = Confidence Interval. P-values < 0.05 indicate statistical significance. Education reference: < 9th grade. Sodium and Protein are Willett energy-adjusted residuals.",
    general_title = "Note: "
  )
```

## 5.1 Forest Plot — Visualisation of Adjusted Odds Ratios

```{r forest_plot, fig.width=8, fig.height=7, fig.cap="Figure 1. Adjusted Odds Ratios with 95% confidence intervals for all predictors of heart failure. Estimates from survey-weighted quasi-binomial logistic regression (NHANES 2011–2018). Dashed vertical line at OR = 1.0 (null hypothesis). X-axis on log scale."}
# Exclude intercept from plot
plot_data <- table_OR %>%
  filter(Variable != "Intercept") %>%
  mutate(
    Variable    = reorder(Variable, OR),
    Significant = ifelse(CI_2.5 > 1 | CI_97.5 < 1, "Yes", "No")
  )

ggplot(plot_data, aes(x = OR, y = Variable, colour = Significant)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50", linewidth = 0.5) +
  geom_errorbarh(aes(xmin = CI_2.5, xmax = CI_97.5), height = 0.25, linewidth = 0.6) +
  geom_point(size = 2.8) +
  scale_x_log10() +
  scale_colour_manual(values = c("Yes" = "#2166AC", "No" = "#888888"),
                      guide  = "none") +
  labs(
    x       = "Odds Ratio (log scale)",
    y       = NULL,
    caption = "Blue = CI excludes 1.0 (statistically significant at α=0.05). Grey = CI crosses 1.0."
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y        = element_text(size = 9)
  )
```

# 6. Model Diagnostics

## 6.1 Discrimination — Area Under the ROC Curve (AUC)

```{r roc_auc, fig.width=5.5, fig.height=5, fig.cap="Figure 2. Receiver Operating Characteristic (ROC) curve for the final logistic model. AUC reflects the model's ability to discriminate between HF and non-HF participants. Note: standard ROC/AUC does not account for complex survey design; this should be interpreted as an approximation."}
# Note: pROC operates on unweighted fitted values. This is a standard
# approximation; survey-weighted AUC estimation requires specialised methods
# (e.g., Lumley & Scott 2015) not yet in base pROC.
roc_obj <- roc(
  response  = analysis_data$HeartFailure,
  predictor = fitted(final_model),
  quiet     = TRUE
)

auc_val <- round(as.numeric(auc(roc_obj)), 3)
cat(sprintf("AUC (unweighted approximation): %.3f\n", auc_val))

plot(roc_obj,
     col  = "#2166AC",
     lwd  = 2,
     main = sprintf("ROC Curve — AUC = %.3f", auc_val),
     print.auc = FALSE)
```

## 6.2 Multicollinearity — Variance Inflation Factors (VIF)

```{r vif, warning=TRUE}
# VIF computed on unweighted lm() as approximation — car::vif() does not
# support svyglm objects directly.
#
# car::vif() returns:
#   - a named numeric VECTOR when all predictors are continuous or binary
#   - a MATRIX with columns (GVIF, Df, GVIF^(1/(2*Df))) when the model
#     contains polytomous factors (e.g. Education with 5 levels,
#     Diabetes with 3 levels, Protein_Origin with 3 levels).
#
# The GVIF^(1/(2*Df)) column is the comparable quantity to a standard VIF:
# squaring it gives an equivalent single-df VIF. Values > 2.24 on this scale
# correspond to VIF > 5; values > 3.16 correspond to VIF > 10.
# Reference: Fox & Monette (1992), JASA.

vif_model <- lm(
  HeartFailure ~ Sodium_Willett + Protein_Willett + Protein_Origin +
    Age + Gender + Poverty_Index_Ratio + Education +
    BMI + Diabetes + Mean_Systolic + Medication_Use,
  data = analysis_data
)

vif_raw <- vif(vif_model)

# Handle both output shapes robustly
if (is.matrix(vif_raw)) {
  # Polytomous factors present: use GVIF^(1/(2*Df)) column
  vif_table <- data.frame(
    Variable  = rownames(vif_raw),
    GVIF_adj  = round(vif_raw[, "GVIF^(1/(2*Df))"], 3),
    Df        = as.integer(vif_raw[, "Df"]),
    row.names = NULL
  ) %>%
    mutate(
      # Square the adjusted GVIF to recover a VIF-equivalent for interpretation
      VIF_equiv = round(GVIF_adj^2, 2),
      Concern   = case_when(
        VIF_equiv > 10 ~ "Severe",
        VIF_equiv > 5  ~ "Moderate",
        TRUE           ~ "Acceptable"
      )
    ) %>%
    arrange(desc(VIF_equiv))

  kable(
    vif_table,
    col.names = c("Variable", "GVIF^(1/(2·Df))", "Df", "VIF equivalent", "Assessment"),
    align     = "lcccc",
    caption   = paste(
      "Table 3. Generalized Variance Inflation Factors (GVIF) for polytomous predictors.",
      "GVIF^(1/(2·Df)) is the multi-df generalisation of sqrt(VIF).",
      "VIF equivalent = [GVIF^(1/(2·Df))]². Values > 5 indicate moderate collinearity."
    )
  ) %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
    footnote(
      general       = "Fox & Monette (1992). Unweighted lm() approximation; interpret directionally.",
      general_title = "Reference: "
    )

} else {
  # All binary / continuous predictors: standard VIF vector
  vif_table <- data.frame(
    Variable  = names(vif_raw),
    VIF       = round(as.numeric(vif_raw), 2),
    row.names = NULL
  ) %>%
    mutate(
      Concern = case_when(
        VIF > 10 ~ "Severe",
        VIF > 5  ~ "Moderate",
        TRUE     ~ "Acceptable"
      )
    ) %>%
    arrange(desc(VIF))

  kable(
    vif_table,
    col.names = c("Variable", "VIF", "Assessment"),
    align     = "lcc",
    caption   = "Table 3. Variance Inflation Factors (unweighted approximation). Values > 5 indicate moderate collinearity."
  ) %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
}
```

# 7. Limitations

The following limitations should be disclosed in the final presentation and any written submission:

1.  **Cross-sectional design:** NHANES is cross-sectional; causality cannot be inferred. The associations between sodium/protein intake and HF prevalence may reflect reverse causation (HF patients may alter their diet after diagnosis).

2.  **Self-reported heart failure:** The outcome (`MCQ160B`) relies on self-reported diagnosis of congestive HF. Misclassification bias is possible. The absence of biomarkers (BNP, troponin) or echocardiography precludes clinical confirmation, though this approach is standard in NHANES-based HF research.

3.  **Single 24-hour dietary recall:** Dietary intake is assessed via a single-day recall, which may not reflect habitual intake. Willett's residual method partially mitigates within-person variability but does not eliminate it.

4.  **Complete Case Analysis:** Participants with missing data were excluded. Given that missingness correlates with socioeconomic status, the analytic sample may over-represent higher-income adults, limiting generalisability to marginalised populations.

5.  **Protein origin classification:** Animal vs. plant protein classification based on the first digit of USDA FNDDS food codes is a coarse approximation. Mixed or processed foods may be systematically misclassified.

6.  **Medication variable scope:** The antihypertensive/diuretic flag is based on active ingredient name matching in `RXDDRUG`. Trade names and combination products not included in the keyword list may result in some participants being incorrectly classified as unexposed.

# 8. Exported Data for Dashboard (Graphical Interface)

The following matrices serve as baselines for the interactive Shiny dashboard.

## 8.1 Population Percentiles (Relative Comparison)

Survey-weighted percentiles of raw sodium and protein intake, used to contextualise user inputs against the representative US population.

```{r percentile}
nutrition_percentiles <- svyquantile(
  ~Sodium_mg + Protein_g,
  design    = nhanes_design,
  quantiles = c(0.05, 0.25, 0.50, 0.75, 0.90, 0.95),
  na.rm     = TRUE
)

table_percentiles <- as.data.frame(
  do.call(rbind, lapply(nutrition_percentiles, function(x) x[, 1]))
)
colnames(table_percentiles)  <- c("P5", "P25", "P50", "P75", "P90", "P95")
rownames(table_percentiles)  <- c("Sodium (mg)", "Protein (g)")

kable(table_percentiles, digits = 1, align = "c",
      caption = "Table 4. Survey-weighted population percentiles for sodium and protein intake (NHANES 2011–2018 adults). Used as reference distribution in the Shiny dashboard.") %>%
  kable_styling(bootstrap_options = c("striped", "condensed", "hover"),
                full_width = FALSE)
```

## 8.2 Predicted Risk Curve Matrix

Predicted HF probability for a reference patient profile across the observed range of Willett-adjusted sodium intake (P5–P95). The sodium range is derived from the data rather than fixed values, ensuring the curve covers clinically meaningful variation.

```{r risk_curve, fig.width=6.5, fig.height=4.5, fig.cap="Figure 3. Predicted probability of heart failure as a function of energy-adjusted sodium intake (Willett residual). Reference profile: Male, age 50, BMI 28, mean protein intake, PIR 2.0, high school education, no diabetes, mean systolic blood pressure, no antihypertensive medication. Shaded band = 95% CI."}

# --- Robust svyquantile scalar extraction ---
# svyquantile can return a named vector, a matrix, or a list depending on the
# survey package version. Using [[1]][1] guarantees a length-1 numeric scalar
# regardless of the return structure.
extract_scalar_quantile <- function(formula, design, prob) {
  q <- svyquantile(formula, design = design, quantiles = prob, na.rm = TRUE)
  as.numeric(q[[1]])[1]
}

sodium_p5  <- extract_scalar_quantile(~Sodium_Willett, nhanes_design, 0.05)
sodium_p95 <- extract_scalar_quantile(~Sodium_Willett, nhanes_design, 0.95)

sodium_scenarios <- data.frame(
  Sodium_Willett      = seq(sodium_p5, sodium_p95, length.out = 50),
  Protein_Willett     = mean(analysis_data$Protein_Willett, na.rm = TRUE),
  Protein_Origin      = factor("Mixed", levels = c("Mixed", "Animal", "Plant")),
  Age                 = 50,
  Gender              = factor("Male", levels = c("Male", "Female")),
  Poverty_Index_Ratio = 2.0,
  Education    = factor("High School Graduate",
                               levels = levels(analysis_data$Education)),
  BMI                 = 28,
  Diabetes            = factor("No", levels = c("No", "Pre-diabetes", "Yes")),
  Mean_Systolic       = mean(analysis_data$Mean_Systolic, na.rm = TRUE),
  Medication_Use      = factor("No", levels = c("No", "Yes"))
)

# Predicted probabilities — svyglm predict() returns an svystat object,
# NOT a list. Fitted values are the object itself; variances are in attr(...,"var").
pred_link <- predict(final_model, newdata = sodium_scenarios, type = "link", se = TRUE)

# Back-transform from log-odds to probability
inv_logit <- function(x) exp(x) / (1 + exp(x))

pred_prob <- inv_logit(as.numeric(pred_link))
pred_se   <- as.numeric(sqrt(diag(attr(pred_link, "var"))))  # SE on the link scale

# CI on link scale first, then back-transform (avoids probabilities outside [0,1])
ci_lower <- inv_logit(as.numeric(pred_link) - 1.96 * pred_se)
ci_upper <- inv_logit(as.numeric(pred_link) + 1.96 * pred_se)

dashboard_matrix <- data.frame(
  Sodium_Willett_mg     = round(sodium_scenarios$Sodium_Willett, 0),
  Estimated_HF_Risk_Pct = round(pred_prob * 100, 2),
  CI_Lower_Pct          = round(ci_lower  * 100, 2),
  CI_Upper_Pct          = round(ci_upper  * 100, 2)
)

# Plot the risk curve
ggplot(dashboard_matrix, aes(x = Sodium_Willett_mg, y = Estimated_HF_Risk_Pct)) +
  geom_ribbon(aes(ymin = CI_Lower_Pct, ymax = CI_Upper_Pct),
              fill = "#2166AC", alpha = 0.15) +
  geom_line(colour = "#2166AC", linewidth = 1) +
  labs(
    x = "Energy-adjusted sodium intake — Willett residual (mg)",
    y = "Estimated HF probability (%)"
  ) +
  theme_minimal(base_size = 10)

# Table export for Shiny (first 10 rows shown)
kable(head(dashboard_matrix, 10),
      col.names = c("Willett sodium (mg)", "HF risk (%)", "CI lower (%)", "CI upper (%)"),
      digits    = 2, align = "cccc",
      caption   = "Table 5. Predicted HF risk matrix (first 10 rows). Full matrix exported as dashboard_matrix for Shiny integration.") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)

# Export full matrix and model for Shiny dashboard
saveRDS(dashboard_matrix, file = "data/dashboard_matrix.rds")
saveRDS(final_model,      file = "data/final_model.rds")
```
